Neon: Nuclear Norm to Beat Muon

### 2025-04-27 article/muon.tex
added missing minus
modified paper a little, to make it obvious that we still need to figure out which version of Neon we accept

Using the kernel norm in \eqref{eqn:opt_problem_mat} leads to a rank-one update of the weight matrices:
\begin{equation}\label{eqn:update_star}
  \delta W = \frac{1}{\lambda} u_1 \sigma_1 v_1^T\,,
\end{equation}
where $\sigma_1$ is the largest singular value, and $u_1$ and $v_1$ are the corresponding singular vectors. We expect one iteration of this method to be significantly faster than one iteration of Muon.

Another choice is the $F*$ norm. With this choice, \eqref{eqn:opt_problem_mat} yields 
\begin{equation}\label{eqn:update_F_star}
\delta W = \frac{1}{\lambda}UDV^T
\end{equation} 
with $D = \text{diag}(d_i)$, where $d_i = [\sigma_i - \tau]_+$ and $\tau$ is given by:
\begin{equation}
    \sum_{i=1}^n [\sigma_i - \tau]_+ = \tau\,.
\end{equation}
We anticipate that the method with this update step will perform well with large batch sizes.

In this article we show how one can quickly compute weight updates defined by \eqref{eqn:update_star} and \eqref{eqn:update_F_star}. Then we finilize the methods by adding momentum and test their performance against those of Muon ant training multilayer perceptron and transformer. The results will be fast algorithms, which we will convert into a new optimizer classes for PyTorch, as was done with Muon.

\section{Outcomes}
The main results of the project are expected to be:
\begin{enumerate}
    \item Deriving of Neon and argumentation of the numerical methods which will be used in the algorithm
    \item Code implementation of Neon as PyTorch optimizer class
    \item Numerical experiments comparing performance of training with Neon, Muon, SVD and Adam on several architectures: MLP, CNN, and NanoGPT.
    \item Participation in CIFAR-10 challenge with Neon (if the alrothims turns out to be competitive)
    \item Article describing all the results. If they are satisfactory, we intend to be in time to apply for NeurIPS 2025. It is a challenge, but I think we must give it a try.
\end{enumerate}

@@

While the uniqie advantages and effects introduced by Neon are yet to be discovered, we can already say that our new optimizer introduces additional overhead of $???$[Depends on variant of Neon we choose] FLOPS per $m \times n$ weight matrix, which is much better then $O(m n \min\{m, n\})$ for Muon optimizer and $O(n^3 + m^3)$ for Shampoo.

@@

\section{Prototyping phase report}
\begin{enumerate}
    \item Update rule is derived, see idea
    \item Update rule methods are tested: power iteration vs Lanzos (see playground)
    \item Recorded the distribution of singular values of gradients during NanoGPT training (add fig).
    \item NanoGPT is tested on Muon and Adam (add fig). The results are strange, because the methods do not converge at the expected rate (Muon is slower). Besides, Neon doe not converge (add fig).
    \item Neon, Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_simple.png}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_simple.png}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
\end{enumerate}

>>>>

Using the kernel norm in \eqref{eqn:opt_problem_mat} leads to a rank-one update of the weight matrices:
\begin{equation}\label{eqn:update_star}
  \delta W = -\frac{1}{\lambda} u_1 \sigma_1 v_1^T\,,
\end{equation}
where $\sigma_1$ is the largest singular value, and $u_1$ and $v_1$ are the corresponding singular vectors. We expect one iteration of this method to be significantly faster than one iteration of Muon.

Another choice is the $F*$ norm. With this choice, \eqref{eqn:opt_problem_mat} yields 
\begin{equation}\label{eqn:update_F_star}
\delta W = -\frac{1}{\lambda}UDV^T
\end{equation} 
with $D = \text{diag}(d_i)$, where $d_i = [\sigma_i - \tau]_+$ and $\tau$ is given by:
\begin{equation}
    \sum_{i=1}^n [\sigma_i - \tau]_+ = \tau\,.
\end{equation}
We anticipate that the method with this update step will perform well with large batch sizes.

In this article we show how one can quickly compute weight updates defined by \eqref{eqn:update_star} or \eqref{eqn:update_F_star} (for now we need more tests to accept one variant or both of them). Then we finilize the method by adding momentum and test their performance against those of Muon at training multilayer perceptron and transformer. The results will be fast algorithm, which we will convert into a new optimizer classes for PyTorch, as was done with Muon.

\section{Outcomes}
The main results of the project are expected to be:
\begin{enumerate}
    \item Deriving of Neon and argumentation of the numerical methods which will be used in the algorithm
    \item Code implementation of Neon as PyTorch optimizer class
    \item Numerical experiments comparing performance of training with Neon, Muon, SVD and Adam on several architectures: MLP, CNN, and NanoGPT.
    \item Participation in CIFAR-10 challenge with Neon (if the alrothim turns out to be competitive)
    \item Article describing all the results. If they are satisfactory, we intend to be in time to apply for NeurIPS 2025. It is a challenge, but I think we must give it a try.
\end{enumerate}

@@

While the uniqie advantages and effects introduced by Neon are yet to be discovered, we can already say that our new optimizer introduces additional overhead of $O(mn)$ FLOPS on average per $m \times n$ weight matrix, which is much better then $O(m n \min\{m, n\})$ for Muon optimizer and $O(n^3 + m^3)$ for Shampoo.

@@

\section{Prototyping phase report}
\begin{enumerate}
    \item Update rule is derived, see idea
    \item Update rule methods are tested: power iteration vs Lanzos (see playground)
    \item Recorded the distribution of singular values of gradients during NanoGPT training (add fig).
    \item NanoGPT is tested on Muon and Adam (add fig). The results are strange, because the methods do not converge at the expected rate (Muon is slower). Besides, Neon (rank-1 version) does not converge (add fig).
    \item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_simple.png}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_simple.png}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
\end{enumerate}
